---
title: "Final Portfolio Project"
author: "Seth Bonney and Cuong Tran"
format: pdf
geometry: margin=0.75in
---

# Introduction
Sleep is one of the first things students sacrifice when school gets busy, but it plays a major role in stress, focus, and academic performance. Late-night studying and inconsistent routines often create a "night-owl" pattern that feels normal - but might carry hidden costs. In this project, we explore how sleep duration relates to sleep quality, stress levels, study habits, and academic outcomes. To do this, we use three real datasets: the Sleep Health and Lifestyle dataset, which includes sleep duration, sleep quality, stress, age, and gender; the Student Performance dataset, which contains performance index, previous scores, study time, and reported sleep hours; and the Student Performance Factors dataset, which adds information on exam scores, study hours, and additional behavioral variables.

By analyzing these datasets individually and together, we look for consistent patterns connecting sleep to academic performance and well-being. Our visualizations, summaries, and interactive Shiny app help us ask whether students who sleep more report lower stress, whether sleep quality improves with longer sleep, and whether better sleep is linked to stronger academic outcomes. Although sleep is not the only factor shaping performance, our results suggest that adequate sleep is associated with better well-being and slightly higher academic achievement. This helped as highlight the importance of healthy sleep habits in a student’s daily life.

# Research Questions
We focused on how sleeping relates to stress and academic performance using three datasets on sleep and student outcomes. Our project talks about three questions. The first one was: What do general sleep patterns look like in the Sleep Health and Lifestyle data? Which entails sub-questions such as How much do people sleep on average? How are sleep duration, sleep quality, and stress levels related? The second question was: How are sleep and academic performance related in student-level data? Which also covers sub-questions like, Is more sleep associated with higher performance index and exam scores? How does sleep compare to study time in predicting performance? And the last question was: How do sleep and study behaviors interact? Which also had sub-questions like Do students who sleep more also study more (or are they more efficient)? Are there "trade-offs" between sleep, stress, and performance?

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(broom)

# Import data
sleep_raw   <- read_csv("Sleep_health_and_lifestyle_dataset.csv")
student_raw <- read_csv("Student_Performance.csv")
factors_raw <- read_csv("StudentPerformanceFactors.csv")
```


```{r}
#| echo: false
#| message: false
#| warning: false
#| 
sleep_clean_name <- sleep_raw |> 
  clean_names() |>
  mutate(across(where(is.character), 
                factor)
         )

student_clean_name <- student_raw |> 
  clean_names() |>
  mutate(across(where(is.character), 
                factor)
         )

factors_clean_name <- factors_raw |> 
  clean_names() |>
  mutate(across(where(is.character), 
                factor)
         )

```


```{r}
#| echo: false
#| message: false
#| warning: false

# Clean Sleep dataset
sleep <- sleep_raw |>
  clean_names() |>
  select(
    sleep_duration,              
    sleep_quality = quality_of_sleep,
    stress_level,
    age,
    gender
  ) |>
  drop_na(sleep_duration, sleep_quality, stress_level)

# Create sleep categories
sleep <- sleep |>
  mutate(
    sleep_category = case_when(
      sleep_duration < 7 ~ "Short sleep",
      sleep_duration >= 7 & sleep_duration < 9 ~ "Recommended sleep",
      sleep_duration >= 9 ~ "Long sleep"
    ),
    sleep_category = factor(
      sleep_category,
      levels = c("Short sleep", "Recommended sleep", "Long sleep")
    )
  )

# Clean Student Performance dataset
student <- student_raw |>
  clean_names()

student <- student |>
  select(
    performance_index,
    previous_scores,
    hours_studied,
    sleep_hours
  ) |>
  drop_na(performance_index, previous_scores, hours_studied)

# Clean Student Performance Factors dataset
factors <- factors_raw |>
  clean_names()
factors <- factors |>
  select(
    study_hours = hours_studied,     
    score = previous_scores,
    sleep_hours = sleep_hours
  ) |>
  drop_na()

# Some summary statistics 
sleep_summary <- sleep |>
  summarise(
    mean_sleep   = mean(sleep_duration),
    median_sleep = median(sleep_duration),
    sd_sleep     = sd(sleep_duration),
    mean_quality = mean(sleep_quality),
    mean_stress  = mean(stress_level)
  )

student_summary <- student |>
  summarise(
    mean_grade   = mean(performance_index),
    median_grade = median(performance_index),
    mean_prev    = mean(previous_scores),
    mean_study   = mean(hours_studied)
  )

factor_summary <- factors |>
  summarise(
    mean_score   = mean(score),
    mean_sleep    = mean(sleep_hours),
    mean_study   = mean(study_hours)
  )
```

```{r}
#| echo: false
#| message: false
#| warning: false
# Data Joining
avg_factors <- factors |>
  group_by(sleep_hours) |>
  summarise(avg_study = mean(study_hours, na.rm = TRUE),
            avg_score = mean(score, na.rm = TRUE))

avg_sleep <- sleep |>
  mutate(sleep_hours = round(sleep_duration)) |>
  group_by(sleep_duration) |>
  summarise(avg_sleep_quality = mean(sleep_quality, na.rm = TRUE),
            avg_stress_level = mean(stress_level, na.rm = TRUE),
            sleep_hours = sleep_hours)

joined_sleep_factors <- left_join(avg_factors, avg_sleep, by = join_by(sleep_hours))

avg_students <- student |>
  group_by(sleep_hours) |>
  summarise(avg_performance = mean(performance_index, na.rm = TRUE),
            avg_previous_score = mean(previous_scores, na.rm = TRUE),
            avg_hours_studied = mean(hours_studied, na.rm = TRUE))

full_data <- full_join(joined_sleep_factors, avg_students, by = join_by(sleep_hours)) |> drop_na() |> distinct()


```

```{r}
#| echo: false
#| message: false
#| warning: false
full_data_clean <- full_data |>
  select(-c(sleep_hours)) |>
  relocate(sleep_duration)

```


# Interpreting Sleep Patterns
```{r}
#| echo: false
#| message: false
#| warning: false
library(ggplot2)
ggplot(sleep, aes(sleep_duration)) +
  geom_histogram(bin = 20) +
  theme_minimal() +
  labs(
    title = "Distribution of Sleep Duration",
    x = "Sleep hours",
    y = "Count"
  )
```


```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(sleep, aes(x = sleep_duration, y = sleep_quality)) +
  geom_jitter(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  theme_minimal() +
  labs(
    title = "Sleep Duration vs Sleep Quality",
    x = "Sleep Duration (hours)",
    y = "Sleep Quality"
  )
```


```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(sleep, aes(x = sleep_duration, y = stress_level)) +
  geom_jitter(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  theme_minimal() +
  labs(
    title = "Sleep Duration vs Stress Level",
    x = "Sleep Duration (hours)",
    y = "Stress Level"
  )
```

# Sleep, Study Time, and Performance (Student Dataset)
```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(student, aes(x = hours_studied, y = performance_index)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "purple") +
  theme_minimal() +
  labs(
    title = "Study Time vs Performance Index",
    x = "Hours Studied",
    y = "Performance Index"
  )
```


```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(student, aes(x = sleep_hours, y = performance_index)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "forestgreen") +
  theme_minimal() +
  labs(
    title = "Sleep Hours vs Academic Performance",
    x = "Sleep Hours",
    y = "Performance Index"
  )
```


```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(student, aes(x = previous_scores, y = performance_index)) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Previous Scores vs Academic Performance",
    x = "Previous Scores",
    y = "Performance Index"
  )
```


# Study Behavior and Exam Scores (Factors Dataset)
```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(factors, aes(x = study_hours, y = score)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  theme_minimal() +
  labs(
    title = "Study Hours vs Exam Score",
    x = "Study Hours",
    y = "Exam Score"
  )
```


```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(factors, aes(x = sleep_hours, y = score)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "navy") +
  theme_minimal() +
  labs(
    title = "Sleep Hours vs Exam Score",
    x = "Sleep Hours",
    y = "Exam Score"
  )
```


# How Sleep Connects to Performance Across Datasets

```{r}
#| echo: false
#| message: false
#| warning: false
# Sleep hours vs academic performance index
ggplot(full_data, aes(x = sleep_duration, y = avg_performance)) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Average Performance Index by Sleep Hours",
    x = "Sleep Hours (rounded)",
    y = "Average Performance Index"
  )
```
```{r}
#| echo: false
#| message: false
#| warning: false
# Sleep hours vs previous scores
ggplot(full_data, aes(x = sleep_duration, y = avg_previous_score)) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Average Previous Scores by Sleep Hours",
    x = "Sleep Hours (rounded)",
    y = "Average Previous Score"
  )
```
```{r}
#| echo: false
#| message: false
#| warning: false
# Sleep hours vs hours studied
ggplot(full_data, aes(x = sleep_duration, y = avg_hours_studied)) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Average Study Time by Sleep Hours",
    x = "Sleep Hours (rounded)",
    y = "Average Hours Studied"
  )
```
```{r}
#| echo: false
#| message: false
#| warning: false
# Sleep hours vs exam score (from factors dataset)
ggplot(full_data, aes(x = sleep_duration, y = avg_score)) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Average Exam Score by Sleep Hours",
    x = "Sleep Hours (rounded)",
    y = "Average Exam Score"
  )
```
```{r}
#| echo: false
#| message: false
#| warning: false
# Sleep hours vs sleep quality
ggplot(full_data, aes(x = sleep_duration, y = avg_sleep_quality)) +
  geom_point() +
  geom_smooth(se = TRUE, color = "navy") +
  theme_minimal() +
  labs(
    title = "Average Sleep Quality by Sleep Hours",
    x = "Sleep Hours (rounded)",
    y = "Average Sleep Quality"
  )
```
```{r}
#| echo: false
#| message: false
#| warning: false
# Sleep hours vs stress level
ggplot(full_data, aes(x = sleep_duration, y = avg_stress_level)) +
  geom_point() +
  geom_smooth(se = TRUE, color = "navy") +
  theme_minimal() +
  labs(
    title = "Average Stress Level by Sleep Hours",
    x = "Sleep Hours (rounded)",
    y = "Average Stress Level"
  )
```
