---
title: "Final Portfolio Project"
author: "Seth Bonney and Cuong Tran"
format: pdf
geometry: margin=0.75in
---

# Research Questions
We investigate how “being a night owl” relates to stress and academic performance using three datasets on sleep and student outcomes. Our project focuses on three questions:

### 1. What do general sleep patterns look like in the Sleep Health and Lifestyle data?
How much do people sleep on average? How are sleep duration, sleep quality, and stress levels related?

### 2. How are sleep and academic performance related in student-level data?
Is more sleep associated with higher performance index and exam scores? How does sleep compare to study time in predicting performance?

### 3. How do sleep and study behaviors interact?
Do students who sleep more also study more (or more efficiently)? Are there “trade-offs” between sleep, stress, and performance?

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(broom)

# Import data
sleep_raw   <- read_csv("Sleep_health_and_lifestyle_dataset.csv")
student_raw <- read_csv("Student_Performance.csv")
factors_raw <- read_csv("StudentPerformanceFactors.csv")
```


```{r}
#| echo: false
#| message: false
#| warning: false
#| 
sleep_clean_name <- sleep_raw |> 
  clean_names() |>
  mutate(across(where(is.character), 
                factor)
         )

student_clean_name <- student_raw |> 
  clean_names() |>
  mutate(across(where(is.character), 
                factor)
         )

factors_clean_name <- factors_raw |> 
  clean_names() |>
  mutate(across(where(is.character), 
                factor)
         )

```


```{r}
#| echo: false
#| message: false
#| warning: false

# Clean Sleep dataset
sleep <- sleep_raw |>
  clean_names() |>
  select(
    sleep_duration,              
    sleep_quality = quality_of_sleep,
    stress_level,
    age,
    gender
  ) |>
  drop_na(sleep_duration, sleep_quality, stress_level)

# Create sleep categories
sleep <- sleep |>
  mutate(
    sleep_category = case_when(
      sleep_duration < 7 ~ "Short sleep",
      sleep_duration >= 7 & sleep_duration < 9 ~ "Recommended sleep",
      sleep_duration >= 9 ~ "Long sleep"
    ),
    sleep_category = factor(
      sleep_category,
      levels = c("Short sleep", "Recommended sleep", "Long sleep")
    )
  )

# Clean Student Performance dataset
student <- student_raw |>
  clean_names()

student <- student |>
  select(
    performance_index,
    previous_scores,
    hours_studied,
    sleep_hours
  ) |>
  drop_na(performance_index, previous_scores, hours_studied)

# Clean Student Performance Factors dataset
factors <- factors_raw |>
  clean_names()
factors <- factors |>
  select(
    study_hours = hours_studied,     
    score = previous_scores,
    sleep_hours = sleep_hours
  ) |>
  drop_na()

# Some summary statistics 
sleep_summary <- sleep |>
  summarise(
    mean_sleep   = mean(sleep_duration),
    median_sleep = median(sleep_duration),
    sd_sleep     = sd(sleep_duration),
    mean_quality = mean(sleep_quality),
    mean_stress  = mean(stress_level)
  )

student_summary <- student |>
  summarise(
    mean_grade   = mean(performance_index),
    median_grade = median(performance_index),
    mean_prev    = mean(previous_scores),
    mean_study   = mean(hours_studied)
  )

factor_summary <- factors |>
  summarise(
    mean_score   = mean(score),
    mean_sleep    = mean(sleep_hours),
    mean_study   = mean(study_hours)
  )

# Data Joining
avg_factors <- factors |>
  group_by(sleep_hours) |>
  summarise(avg_study = mean(study_hours, na.rm = TRUE),
            avg_score = mean(score, na.rm = TRUE))

avg_sleep <- sleep |>
  mutate(sleep_hours = round(sleep_duration)) |>
  group_by(sleep_duration) |>
  summarise(avg_sleep_quality = mean(sleep_quality, na.rm = TRUE),
            avg_stress_level = mean(stress_level, na.rm = TRUE),
            sleep_hours = sleep_hours)

joined_sleep_factors <- left_join(avg_factors, avg_sleep, by = join_by(sleep_hours))

avg_students <- student |>
  group_by(sleep_hours) |>
  summarise(avg_performance = mean(performance_index, na.rm = TRUE),
            avg_previous_score = mean(previous_scores, na.rm = TRUE),
            avg_hours_studied = mean(hours_studied, na.rm = TRUE))

full_data <- full_join(joined_sleep_factors, avg_students, by = join_by(sleep_hours)) |> drop_na() |> distinct()


```

```{r}
#| echo: false
#| message: false
#| warning: false
full_data_clean <- full_data |>
  select(-c(sleep_hours)) |>
  relocate(sleep_duration)

```


# Interpreting Sleep Patterns
```{r}
#| echo: false
#| message: false
#| warning: false
library(ggplot2)
ggplot(sleep, aes(sleep_duration)) +
  geom_histogram(bins = 20) +
  theme_minimal() +
  labs(
    title = "Distribution of Sleep Duration",
    x = "Sleep hours",
    y = "Count"
  )
```
The histogram of sleep duration shows that most individuals in the Sleep Health and Lifestyle dataset sleep between 6 and 8 hours per night, with a fairly tight distribution around this range. Very short or very long sleep is rare. 


```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(sleep, aes(sleep_category)) +
  geom_bar(fill = "skyblue", alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "Distribution of Sleep Categories",
    x = "Sleep Category",
    y = "Count"
  )
```
The bar chart of sleep categories highlights that the majority of observations fall into the “Recommended sleep” (7–9 hours) group, with a smaller but non-trivial share of “Short sleep” cases (under 7 hours). 



```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(sleep, aes(x = sleep_duration, y = sleep_quality)) +
  geom_jitter(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  theme_minimal() +
  labs(
    title = "Sleep Duration vs Sleep Quality",
    x = "Sleep Duration (hours)",
    y = "Sleep Quality"
  )
```
The scatterplot of sleep duration vs. sleep quality shows a clear positive relationship: as sleep duration increases, average sleep quality rises. The regression line slopes upward, suggesting that, on average, sleeping more is associated with better-reported sleep quality.


```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(sleep, aes(x = sleep_duration, y = stress_level)) +
  geom_jitter(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  theme_minimal() +
  labs(
    title = "Sleep Duration vs Stress Level",
    x = "Sleep Duration (hours)",
    y = "Stress Level"
  )
```
In contrast, sleep duration vs. stress level displays a negative relationship: the regression line slopes downward, indicating that individuals who sleep less tend to report higher stress.

```{r}
#| echo: false
#| message: false
#| warning: false
sleep |>
  group_by(sleep_category) |>
  summarise(mean_stress = mean(stress_level)) |>
  ggplot(aes(sleep_category, mean_stress, fill = sleep_category)) +
  geom_col(alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "Average Stress Level by Sleep Category",
    x = "Sleep Category",
    y = "Mean Stress Level"
  )

```
This pattern is reinforced in the bar chart of mean stress by sleep category, where the Short sleep group has the highest average stress, and stress is lowest among those with recommended sleep. Together, these results show that short sleep is linked to worse sleep quality and higher stress, while recommended sleep is associated with better outcomes on both measures.


# Sleep, Study Time, and Performance (Student Dataset)
```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(student, aes(x = hours_studied, y = performance_index)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "purple") +
  theme_minimal() +
  labs(
    title = "Study Time vs Performance Index",
    x = "Hours Studied",
    y = "Performance Index"
  )
```
In the student performance dataset, we see a modest positive relationship between hours studied and performance index. As hours studied increase, the regression line rises, indicating that students who study more tend to achieve slightly higher performance scores, although there is substantial variability at each study level.

```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(student, aes(x = sleep_hours, y = performance_index)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "forestgreen") +
  theme_minimal() +
  labs(
    title = "Sleep Hours vs Academic Performance",
    x = "Sleep Hours",
    y = "Performance Index"
  )
```
The plot of sleep hours vs. performance index shows a very slight positive slope. Performance does not dramatically jump for any particular sleep value, but students with more sleep tend to have marginally higher performance. This suggests that sleep is beneficial but not the only driver of grades.

```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(student, aes(x = previous_scores, y = performance_index)) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    title = "Previous Scores vs Academic Performance",
    x = "Previous Scores",
    y = "Performance Index"
  )
```
When we examine previous scores vs. performance index, higher previous scores are associated with higher current performance. This is consistent with the idea that prior academic preparation and ability play a large role in current outcomes. Overall, this dataset suggests that both study time and prior performance are important predictors of academic outcomes, while sleep hours have a small but favorable association with performance.

# Study Behavior and Exam Scores (Factors Dataset)
```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(factors, aes(x = study_hours, y = score)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  theme_minimal() +
  labs(
    title = "Study Hours vs Exam Score",
    x = "Study Hours",
    y = "Exam Score"
  )
```
In the Student Performance Factors dataset, study hours vs. exam score shows a generally positive association: as study hours increase, the regression line slopes upward. The effect is not huge, and scores vary a lot at each study level, but the pattern is consistent with the idea that more study is weakly linked to better exam performance.

```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(factors, aes(x = sleep_hours, y = score)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "navy") +
  theme_minimal() +
  labs(
    title = "Sleep Hours vs Exam Score",
    x = "Sleep Hours",
    y = "Exam Score"
  )
```
The sleep hours vs. exam score plot shows a nearly flat line with a very small slope. Exam scores do not change dramatically across the observed range of sleep hours. This suggests that, in this particular dataset, study time has a more obvious direct relationship with scores than sleep does, although sleep might still matter indirectly through stress, focus, or consistency.

# How Sleep Connects to Performance Across Datasets

```{r}
#| echo: false
#| message: false
#| warning: false
# Sleep hours vs academic performance index
ggplot(full_data, aes(x = sleep_duration, y = avg_performance)) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Average Performance Index by Sleep Hours",
    x = "Sleep Hours (rounded)",
    y = "Average Performance Index"
  )

# Sleep hours vs previous scores
ggplot(full_data, aes(x = sleep_duration, y = avg_previous_score)) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Average Previous Scores by Sleep Hours",
    x = "Sleep Hours (rounded)",
    y = "Average Previous Score"
  )

# Sleep hours vs hours studied
ggplot(full_data, aes(x = sleep_duration, y = avg_hours_studied)) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Average Study Time by Sleep Hours",
    x = "Sleep Hours (rounded)",
    y = "Average Hours Studied"
  )

# Sleep hours vs exam score (from factors dataset)
ggplot(full_data, aes(x = sleep_duration, y = avg_score)) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Average Exam Score by Sleep Hours",
    x = "Sleep Hours (rounded)",
    y = "Average Exam Score"
  )

# Sleep hours vs sleep quality
ggplot(full_data, aes(x = sleep_duration, y = avg_sleep_quality)) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Average Sleep Quality by Sleep Hours",
    x = "Sleep Hours (rounded)",
    y = "Average Sleep Quality"
  )

# Sleep hours vs stress level
ggplot(full_data, aes(x = sleep_duration, y = avg_stress_level)) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Average Stress Level by Sleep Hours",
    x = "Sleep Hours (rounded)",
    y = "Average Stress Level"
  )
```

To link the three datasets, we aggregated each one by rounded sleep hours and then joined the summaries. For each sleep-hour value, we calculated:

-   Average performance index and previous scores (student dataset)
-   Average exam score and study hours (factors dataset)
-   Average sleep quality and stress level (sleep dataset)

Across these combined summaries, several patterns emerge:

-   Average performance index is slightly higher for students sleeping closer to the recommended 7–8 hours. The differences are not huge, but there is no evidence that very short sleep improves performance.
-   Average previous scores and average study hours do not change dramatically across sleep levels, reinforcing that study habits and prior preparation are relatively stable, while sleep is an additional layer on top of those behaviors.
-   Average exam scores vary only modestly with sleep hours, again showing small but generally positive effects of sufficient sleep.
-   On the sleep side, average sleep quality rises and average stress level falls as sleep hours increase from short to recommended levels.

Putting these results together, the combined view suggests that sleep is not a magic bullet that determines academic outcomes by itself, but students who sleep within the recommended range tend to experience better sleep quality, lower stress, and slightly better academic performance than short sleepers. This supports the idea that being a “night owl” with consistently short sleep may come with subtle but meaningful academic and well-being costs.
